Interpolation and extrapolation-based performance measurement of Analytical and Artificial Neural Network models in predicting hot deformation behavior of a medium carbon steel

### 2023-01-23 Article.tex
Update latex and Bibliography files

\author[LGP]{Pierre Tize Mha}
\author[ETS]{Prashant Dhondapure}
\author[LGP]{Olivier Pantalé\corref{cor1}}
\author[ETS]{Mohammad Jahazi}
\author[LGP]{Amèvi Tongne}
\ead{Olivier.Pantale@enit.fr}
\ead[url]{http://www.enit.fr}

@@

Considering the large size of blocks, a purely experimental approach, based on trial and error, can not be used by industry, and therefore, reliable predictive tools, such a Finite Element analysis (FEA) codes, have been developed, and are commercially available.
However, the prediction reliability of such analyses is a function of the accuracy of the material constitutive model, which describes the mutual interactions between the strain, the strain rate, and the temperature during deformation.
As mentioned above, precipitation and phase changes are considered negligible as deformation takes place in the austenite at temperatures above the dissolution temperature of most carbides.
Hence, most constitutive models depend on macroscopic parameters which influence the hardening and softening of the material.
A large number of phenomenological, semi-empirical, or physical models \cite{jia2022thermo, costa2016study, rudnytskyj2020constitutive, Pantale-2021} have been developed.
Among these, the Johnson-Cook (JC) \cite{jia2021modified}, Hansel-Spittel (HS) \cite{chadha2018approach}, and Zerilli-Armstrong (ZA) \cite{sim2022modified} models are the best known and most widely available in FEA codes.
Despite their simplicity, each of them suffers from certain shortcomings: as reported by Jia \eal \cite{Jia-2021}, the JC model suffers from a lack of non-coupling between the strain, the strain rate and the temperature, while the HS model is better adapted for higher strain rate conditions \cite{chadha2018approach}.
To circumvent these shortcomings, several modified model forms have been developed \cite{chadha2018approach, liu2021modified, jia2022thermo1, bai2022comparison, zhu2022constitutive}.
However, as reported by many authors \cite{Li-2013, Zhang-2015, Zhou-2019}, even after adjusting the constants, the high temperature flow behavior, particularly when dynamic recrystallization takes place, cannot be accurately predicted, and none of the models is able to accurately predict the flow behavior outside the experimental testing interval that was used to determine the model constants.
Due to its more physics-based formulation, the ZA model and its modified form, MZA \cite{ovesy2020explicit, niu2020constitutive, Muralli-2017, Cheng-2021, Muralli-2021}, and the Arrhenius-type hyperbolic sine constitutive models are preferred to the JC and HS models both for the prediction of the hot deformation behavior and microstructure analysis of the material \cite{derazkola2021review, wang2022deformation, miao2022deformation}.
The Arrhenius formulation has been revised repeatedly to achieve a more accurate determination of the activation energy for high temperature deformation \cite{rudnytskyj2022investigating, ji2020optimization}.
To overcome the strong dependency of the models on specific alloy types, Tize Mha \eal \cite{TizeMha-2022} recently proposed a constitutive model, PTM, whose formulation is independent of the alloy type.
This formulation is based on the MZA model and the polynomial functions of undefined order that are used during the identification.
However, in the PTM model, the high order of the polynomial function (up-to $10$ in some cases) can affect its accuracy.

@@

The JC model, as mentioned above, is one of the most widely used analytical models because it can be applied to many materials under different conditions of strain $\varepsilon$, strain rate $\mdot\varepsilon$, and temperature $T$.
However, the formulation of this model does not take into account the simultaneous effect of strain, strain rate and temperature.
Indeed, it is formulated by describing the effect of each physical parameter ($\varepsilon$, $\mdot\varepsilon$ and $T$) separately as a factor in the mathematical expression of the model, hence its inability to describe the phenomenon of softening induced by temperature.
The equation that describes this model is given as follows \cite{jia2021modified}:
\begin{equation}
\sigma^y=\left(A+B\varepsilon^{p^{n}}\right) \left[1+C\ln\left(\frac{\mdot\varepsilon}{\mdot\varepsilon_0}\right)\right] \left[1-\left(\frac{T-T_0}{T_m-T_0}\right)^{m}\right],\label{eq:Johnson-Cook}
\end{equation}
where $\sigma^y$ is the flow stress, $\varepsilon^p$ is the plastic strain, $A$ is the initial elastic limit of the material, $B$ is the strain hardening coefficient, $n$ is the strain hardening exponent, $C$ and $m$ are the material constants that describe the strain rate hardening coefficient and the thermal softening coefficient, respectively.
The other values are reference values: $\mdot\varepsilon_0$ is the reference strain rate, $T_m$ and thus $T_0$ are the melting temperature ($1460\celsius$ in our case) and the reference temperature, respectively.
For the determination of the parameters of the analytical models, the reference values for strain rate and temperature are $\mdot\varepsilon_0=0.001~\ps$ and $T_0=1050\celsius$, respectively.

@@

The accuracy and predictive ability of the models are usually assessed through certain coefficients such as the Mean Absolute Relative Error ($\MARE$) defined by Equation (\ref{eq:AARE}):
\begin{equation}
\MARE(\%) = \frac{1}{N} \sum_{i=1}^{N}{\left|\frac{\sigma_i^p -\sigma_i^e}{\sigma_i^e}\right|} \times 100, \label{eq:AARE}
\end{equation}
and the Root Mean Square Error ($\RMSE$) defined by Equation (\ref{eq:RMSE}):
\begin{equation}
\RMSE (\MPa) = \sqrt{\frac{1}{N} \sum_{i=1}^{N} \left(\sigma_i^p - \sigma_i^e\right)^2}, \label{eq:RMSE}
\end{equation}
where $\sigma_i^e$ is the experimental value, $\sigma_i^p$ is the value predicted using the given model of the flow stress $\sigma^y$ and $N$ is the total number of data points used to compute those coefficients (in our case $N=21~000$).
For the JC model, $\MARE=14.05~\%$ and $\RMSE=12.00~\MPa$.
As reported by Phaniraj \cite{Phaniraj-2003},  the correlation coefficient ($\R$) is not always an accurate measure to evaluate the reliability of the constitutive law espacially in the case of a highly non-linear functions because it only shows the correlation of the model with respect to the data and not its accuracy. Therefore, a good (high) value of $\R$ (close to $1$) does not necessarily mean a good prediction of the model, but simply establishes a good linearity correlation between the experiment and the prediction, we avoided its use in our analysis.

@@

The Hansel-Spittel model \cite{savaedi2022review} is one of the least known models in terms of integration in FEM codes, although its parameters can more easily be determined than those of the JC or MZA models.
Programming a simple identification script is sufficient to identify its parameters for a given material.
The equation of the HS model is given by the following relation:
\begin{equation}
\sigma^y = A~\e{m_1T}~\varepsilon^{m_2}~\mdot\varepsilon^{m_3}~\e{({m_4}/{\varepsilon})}~(1+\varepsilon)^{m_5T}~\e{m_6\varepsilon}~\mdot\varepsilon^{m_7T}~T^{m_8}, \label{eq:HS}
\end{equation}
where again, $\sigma^y$ is the flow stress, $\varepsilon$ is the strain, $\mdot\varepsilon$ is the strain rate and $T$ is the temperature, as proposed earlier.
The coefficients $A$ and $m_i$ are the 9 parameters of the model to be identified.
However, this model has some shortcomings, notably related to the fact that its accuracy varies according to the number of parameters taken into account during the identification.
For its identification, several authors restrict its expression to a reduced number of only $5$ or $6$ $m_i$ parameters, by forcing a zero value for the other parameters \cite{Chadha-2018, Mehtedi-2015, Rudnytskyj-2020}.

@@

The Arrhenius type model \cite{zhou2022hot} is one of the most used models in the framework of material forming, especially when it comes to studying the material microstructure.
The model takes into account the physical phenomena describing the behavior of the material in the formulation of the relationships between the stress $\sigma^y$, the strain $\varepsilon$, the strain rate $\mdot\varepsilon$ and the temperature $T$ expressed as power law, exponential law and hyperbolic sine.
This makes it easier to describe the softening phenomenon observed in the material due to increasing temperature.
The following equations describe the Arrhenius model:
\begin{equation}
\mdot\varepsilon =
\begin{cases}
A_1 \sigma^{y^{n_1}} \exp{\left(-\frac{Q}{RT}\right)} & \alpha\sigma^y < 0.8\\
A_2 \exp(\beta\sigma^y) \exp{\left(-\frac{Q}{RT}\right)} & \alpha\sigma^y > 1.2\\
A_3 \left[\sinh{\left(\alpha\sigma^y\right)}\right]^{n_2} \exp{\left(-\frac{Q}{RT}\right)} & \text{for all }\sigma^y
\end{cases}
\label{eq:ArDef}
\end{equation}
with:
\begin{equation}
Z = \mdot\varepsilon \exp{\left(\frac{Q}{RT}\right)}, \label{eq:ArZ}
\end{equation}
where $Z$ is the Zenner-Hollomon parameter \cite{liu2019effect}, $\mdot\varepsilon$ is the strain rate ($\ps$), $Q$ is the apparent activation energy ($\text{J~mol}^{-1}$), $R$ is the universal gas constant ($8.314~\text{J~mol}^{-1} \text{K}^{-1}$), $T$ is the absolute temperature (K), $\sigma^y$ is the flow stress (MPa) for a given strain, strain rate and temperature and $A_1$, $A_2$, $A_3$, $n_1$, $n_2$, $\alpha$ and $\beta=\alpha~n_1$ are dependent on the material.
The corresponding values are independent of the temperature and are obtained from the stress-strain curves at different strain rates and temperatures by the regression method.
Combining Equations (\ref{eq:ArDef}) and (\ref{eq:ArZ}) allows to express the flow stress $\sigma^y$ as a function of the $Z$ parameter:
\begin{equation}
\sigma^y = \frac{1}{\alpha} \ln\left\{\left(\frac{Z}{A}\right)^{1/n} + \left[1 + \left(\frac{Z}{A}\right)^{2/n}\right]^{1/2}\right\}
\end{equation}

@@

Experimental tests were performed on a Gleeble thermomechanical simulator for a modified medium carbon alloy to investigate the applicability and predictive accuracy of 5 analytical models and an Artificial Neural Network model over a range of strains ($0.0-0.7$), strain rates ($0.001~\ps-5~\ps$) and temperatures ($1050\celsius-1250\celsius$).
The analytical models selected for this study were the Johnson-Cook (JC) model \cite{jia2021modified}, the Modified Zerilli-Armstrong (MZA) model \cite{Samantaray-2009}, the Hansel-Spittle (HS) model \cite{savaedi2022review}, the Arrhenius (AR) model \cite{zhou2022hot} and the PTM model \cite{TizeMha-2022}.
The ANN model selected is the one introduced by Pantalé \eal \cite{Pantale-2021}.
An analysis of the data from the Gleeble trials and a comparison of the 6 models proposed in this study against the experiments led to the following conclusions.

>>>>

\author[LGP]{Pierre Tize Mha}
\author[ETS]{Prashant Dhondapure}
\author[ETS]{Mohammad Jahazi}
\author[LGP]{Amèvi Tongne}
\author[LGP]{Olivier Pantalé\corref{cor1}}
\ead{Olivier.Pantale@enit.fr}
\ead[url]{http://www.enit.fr}

@@

Considering the large size of blocks, a purely experimental approach, based on trial and error, can not be used by industry, and therefore, reliable predictive tools, such a Finite Element analysis (FEA) codes, have been developed, and are commercially available.
However, the prediction reliability of such analyses is a function of the accuracy of the material constitutive model, which describes the mutual interactions between the strain, the strain rate, and the temperature during deformation.
As mentioned above, precipitation and phase changes are considered negligible as deformation takes place in the austenite at temperatures above the dissolution temperature of most carbides.
Hence, most constitutive models depend on macroscopic parameters which influence the hardening and softening of the material.
A large number of phenomenological, semi-empirical, or physical models \cite{jia2022thermo, costa2016study, rudnytskyj2020constitutive, Pantale-2021} have been developed.
Among these, the Johnson-Cook (JC) \cite{Johnson-1983}, Hansel-Spittel (HS) \cite{chadha2018approach}, and Zerilli-Armstrong (ZA) \cite{Zerilli-1987} models are the best known and most widely available in FEA codes.
Despite their simplicity, each of them suffers from certain shortcomings: as reported by Jia \eal \cite{Jia-2021}, the JC model suffers from a lack of non-coupling between the strain, the strain rate and the temperature, while the HS model is better adapted for higher strain rate conditions \cite{chadha2018approach}.
To circumvent these shortcomings, several modified model forms have been developed \cite{chadha2018approach, liu2021modified, jia2022thermo1, bai2022comparison, zhu2022constitutive, jia2021modified, sim2022modified}.
However, as reported by many authors \cite{Li-2013, Zhang-2015, Zhou-2019}, even after adjusting the constants, the high temperature flow behavior, particularly when dynamic recrystallization takes place, cannot be accurately predicted, and none of the models is able to accurately predict the flow behavior outside the experimental testing interval that was used to determine the model constants.
Due to its more physics-based formulation, the ZA model and its modified form, MZA \cite{ovesy2020explicit, niu2020constitutive, Muralli-2017, Cheng-2021, Muralli-2021}, and the Arrhenius-type hyperbolic sine constitutive models are preferred to the JC and HS models both for the prediction of the hot deformation behavior and microstructure analysis of the material \cite{derazkola2021review, wang2022deformation, miao2022deformation}.
The Arrhenius formulation has been revised repeatedly to achieve a more accurate determination of the activation energy for high temperature deformation \cite{rudnytskyj2022investigating, ji2020optimization}.
To overcome the strong dependency of the models on specific alloy types, Tize Mha \eal \cite{TizeMha-2022} recently proposed a constitutive model, PTM, whose formulation is independent of the alloy type.
This formulation is based on the MZA model and the polynomial functions of undefined order that are used during the identification.
However, in the PTM model, the high order of the polynomial function (up-to $10$ in some cases) can affect its accuracy.

@@

The JC model, as mentioned above, is one of the most widely used analytical models because it can be applied to many materials under different conditions of strain $\varepsilon$, strain rate $\mdot\varepsilon$, and temperature $T$.
However, the formulation of this model does not take into account the simultaneous effect of strain, strain rate and temperature.
Indeed, it is formulated by describing the effect of each physical parameter ($\varepsilon$, $\mdot\varepsilon$ and $T$) separately as a factor in the mathematical expression of the model, hence its inability to describe the phenomenon of softening induced by temperature.
The equation that describes this model is given as follows \cite{Johnson-1983}:
\begin{equation}
\sigma^y=\left(A+B\varepsilon^{p^{n}}\right) \left[1+C\ln\left(\frac{\mdot\varepsilon}{\mdot\varepsilon_0}\right)\right] \left[1-\left(\frac{T-T_0}{T_m-T_0}\right)^{m}\right],\label{eq:Johnson-Cook}
\end{equation}
where $\sigma^y$ is the flow stress, $\varepsilon^p$ is the plastic strain, $A$ is the initial elastic limit of the material, $B$ is the strain hardening coefficient, $n$ is the strain hardening exponent, $C$ and $m$ are the material constants that describe the strain rate hardening coefficient and the thermal softening coefficient, respectively.
The other values are reference values: $\mdot\varepsilon_0$ is the reference strain rate, $T_m$ and thus $T_0$ are the melting temperature ($1460\celsius$ in our case) and the reference temperature, respectively.
For the determination of the parameters of the analytical models, the reference values for strain rate and temperature are $\mdot\varepsilon_0=0.001~\ps$ and $T_0=1050\celsius$, respectively.

@@

The accuracy and predictive ability of the models are usually assessed through certain coefficients such as the Mean Absolute Relative Error ($\MARE$) defined by Equation (\ref{eq:AARE}):
\begin{equation}
\MARE(\%) = \frac{1}{N} \sum_{i=1}^{N}{\left|\frac{\sigma_i^p -\sigma_i^e}{\sigma_i^e}\right|} \times 100, \label{eq:AARE}
\end{equation}
and the Root Mean Square Error ($\RMSE$) defined by Equation (\ref{eq:RMSE}):
\begin{equation}
\RMSE (\MPa) = \sqrt{\frac{1}{N} \sum_{i=1}^{N} \left(\sigma_i^p - \sigma_i^e\right)^2}, \label{eq:RMSE}
\end{equation}
where $\sigma_i^e$ is the experimental value, $\sigma_i^p$ is the value predicted using the given model of the flow stress $\sigma^y$ and $N$ is the total number of data points used to compute those coefficients (in our case $N=21~000$).
For the JC model, $\MARE=14.05~\%$ and $\RMSE=12.00~\MPa$.
As reported by Phaniraj \cite{Phaniraj-2003},  the correlation coefficient ($\R$) is not always an accurate measure to evaluate the reliability of the constitutive law espacially in the case of a highly non-linear functions because it only shows the correlation of the model with respect to the data and not its accuracy.
A good (high) value of $\R$ (close to $1$) does not necessarily mean a good prediction of the model, but simply establishes a good linearity correlation between the experiment and the prediction, we therefore avoided its use in our analysis.

@@

The Hansel-Spittel model \cite{Hensel-1978} is one of the least known models in terms of integration in FEM codes, although its parameters can more easily be determined than those of the JC or MZA models.
Programming a simple identification script is sufficient to identify its parameters for a given material.
The equation of the HS model is given by the following relation:
\begin{equation}
\sigma^y = A~\e{m_1T}~\varepsilon^{m_2}~\mdot\varepsilon^{m_3}~\e{({m_4}/{\varepsilon})}~(1+\varepsilon)^{m_5T}~\e{m_6\varepsilon}~\mdot\varepsilon^{m_7T}~T^{m_8}, \label{eq:HS}
\end{equation}
where again, $\sigma^y$ is the flow stress, $\varepsilon$ is the strain, $\mdot\varepsilon$ is the strain rate and $T$ is the temperature, as proposed earlier.
The coefficients $A$ and $m_i$ are the 9 parameters of the model to be identified.
However, this model has some shortcomings, notably related to the fact that its accuracy varies according to the number of parameters taken into account during the identification.
For its identification, several authors restrict its expression to a reduced number of only $5$ or $6$ $m_i$ parameters, by forcing a zero value for the other parameters \cite{Chadha-2018, Mehtedi-2015, Rudnytskyj-2020}.

@@

The Arrhenius type model \cite{Sellars-1966} is one of the most used models in the framework of material forming, especially when it comes to studying the material microstructure.
The model takes into account the physical phenomena describing the behavior of the material in the formulation of the relationships between the stress $\sigma^y$, the strain $\varepsilon$, the strain rate $\mdot\varepsilon$ and the temperature $T$ expressed as power law, exponential law and hyperbolic sine.
This makes it easier to describe the softening phenomenon observed in the material due to increasing temperature.
The following equations describe the Arrhenius model:
\begin{equation}
\mdot\varepsilon =
\begin{cases}
A_1 \sigma^{y^{n_1}} \exp{\left(-\frac{Q}{RT}\right)} & \alpha\sigma^y < 0.8\\
A_2 \exp(\beta\sigma^y) \exp{\left(-\frac{Q}{RT}\right)} & \alpha\sigma^y > 1.2\\
A_3 \left[\sinh{\left(\alpha\sigma^y\right)}\right]^{n_2} \exp{\left(-\frac{Q}{RT}\right)} & \text{for all }\sigma^y
\end{cases}
\label{eq:ArDef}
\end{equation}
with:
\begin{equation}
Z = \mdot\varepsilon \exp{\left(\frac{Q}{RT}\right)}, \label{eq:ArZ}
\end{equation}
where $Z$ is the Zenner-Hollomon parameter \cite{Zener-1944}, $\mdot\varepsilon$ is the strain rate ($\ps$), $Q$ is the apparent activation energy ($\text{J~mol}^{-1}$), $R$ is the universal gas constant ($8.314~\text{J~mol}^{-1} \text{K}^{-1}$), $T$ is the absolute temperature (K), $\sigma^y$ is the flow stress (MPa) for a given strain, strain rate and temperature and $A_1$, $A_2$, $A_3$, $n_1$, $n_2$, $\alpha$ and $\beta=\alpha~n_1$ are dependent on the material.
The corresponding values are independent of the temperature and are obtained from the stress-strain curves at different strain rates and temperatures by the regression method.
Combining Equations (\ref{eq:ArDef}) and (\ref{eq:ArZ}) allows to express the flow stress $\sigma^y$ as a function of the $Z$ parameter:
\begin{equation}
\sigma^y = \frac{1}{\alpha} \ln\left\{\left(\frac{Z}{A}\right)^{1/n} + \left[1 + \left(\frac{Z}{A}\right)^{2/n}\right]^{1/2}\right\}
\end{equation}

@@

Experimental tests were performed on a Gleeble thermomechanical simulator for a modified medium carbon alloy to investigate the applicability and predictive accuracy of 5 analytical models and an Artificial Neural Network model over a range of strains ($0.0-0.7$), strain rates ($0.001~\ps-5~\ps$) and temperatures ($1050\celsius-1250\celsius$).
The analytical models selected for this study were the Johnson-Cook (JC) model \cite{Johnson-1983}, the Modified Zerilli-Armstrong (MZA) model \cite{Samantaray-2009}, the Hansel-Spittle (HS) model \cite{Hensel-1978}, the Arrhenius (AR) model \cite{Sellars-1966} and the PTM model \cite{TizeMha-2022}.
The ANN model selected is the one introduced by Pantalé \eal \cite{Pantale-2021}.
An analysis of the data from the Gleeble trials and a comparison of the 6 models proposed in this study against the experiments led to the following conclusions.
